---
title: "exercise4"
output: html_document
date: "2023-10-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here we import the library and the dataset used: the Tetragounal dataset gives genetic data for 236 bees from Australia and Southeast Asia. 

```{r}
library(cluster)
library(prabclus)
library(smacof)
library(fpc)
library(rgl)
```

```{r}
data(tetragonula)
```

Now, by using function implemented in the prablcus library we convert the genetic data in a proper way.

```{r}
ta <- alleleconvert(strmatrix=tetragonula)
tai <- alleleinit(allelematrix=ta)
distmatrix <- as.dist(tai$distmat)
```

#MDS plot

Let's see the MDS plot of the distance matrix previously computed.
```{r}
plot(mds(distmatrix))
```

From this plot is still not clear the clustering structure present in the dataset.

#Hierarchical clustering.

Let's apply the hiearachical clustering procedures with single, average and complete methods.

```{r}
method <- c("single", "average", "complete")

for (i in method){
  result_cluster <- hclust(distmatrix ,method = i)
  plot(result_cluster)
}
```

From those plot is not clear which is the best method among those.


Now let's try to find the optimal number of clusters using the Silhouette Average Width

#### Silhouette scores

# Silhouette for the average method
```{r}
asw_average <- NA
average_clusk <- list()
average_sil <- list()

# Look at K between 2 and 30:
for (k in 2:30){
# Hierarchical-average clustering:
  average_clusk[[k]] <- cutree(hclust(distmatrix, method = "average"), k = k)
  
# Computation of silhouettes:
  average_sil[[k]] <- silhouette(average_clusk[[k]],dist=distmatrix)
  
# ASW needs to be extracted:
  asw_average[k] <- summary(average_sil[[k]])$avg.width
  
}
# Plot the ASW-values against K:
plot(1:30,asw_average,type="l",xlab="Number of opt. clusters with average method",ylab="ASW")
plot(average_sil[[which.max(asw_average)]])

```

# Silhouette for the complete method
```{r}
complete_asw <- NA
complete_clusk <- list()
complete_sil <- list()

# Look at K between 2 and 30:
for (k in 2:30){
# Hierarchical- complete clustering:
  complete_clusk[[k]] <- cutree(hclust(distmatrix, method = "complete"), k = k)
  
# Computation of silhouettes:
  complete_sil[[k]] <- silhouette(complete_clusk[[k]],dist=distmatrix)
  
# ASW needs to be extracted:
  complete_asw[k] <- summary(complete_sil[[k]])$avg.width
  
}
# Plot the ASW-values against K:
plot(1:30,complete_asw,type="l",xlab="Number of opt. clusters with complete method",ylab="ASW")
plot(complete_sil[[which.max(complete_asw)]])
```
# Silhouette single
```{r}
single_asw <- NA
single_clusk <- list()
single_sil <- list()

# Look at K between 2 and 30:
for (k in 2:30){
# Hierarchical-single clustering:
  single_clusk[[k]] <- cutree(hclust(distmatrix, method = "single"), k = k)
  
# Computation of silhouettes:
  single_sil[[k]] <- silhouette(single_clusk[[k]],dist=distmatrix)
  
# ASW needs to be extracted:
  single_asw[k] <- summary(single_sil[[k]])$avg.width
  
}
# Plot the ASW-values against K:
plot(1:30,single_asw,type="l",xlab="Number of opt. clusters with single method",ylab="ASW")

plot(single_sil[[which.max(single_asw)]])
```
## The optimal number of clusters when we compute the average and complete method is a number between 10 and 12. 

Let's visualize the clustering in a mds plot and then decide by a visual inspection which will be the optimal number of cluster.

#Complete method, visual inspection of the mds plot when the optimal number of clusters is 10, 11 and 12.
```{r}
for (i in 10:12){
  complete_cut <- cutree(hclust(distmatrix, method = "complete"), k = i)
  plot(mds(distmatrix)$conf, col = complete_cut, pch = clusym[complete_cut])
}
```
Let's do the same thing for the average method.
```{r}
for (i in 10:12){
  average_cut <- cutree(hclust(distmatrix, method = "average"), k = i)
  plot(mds(distmatrix)$conf, col = average_cut, pch = clusym[average_cut])
}

```
The result are pretty similar, the optimal number of clusters is some number between 10 and 12. If we rely to the silhouette I would say that 10 is the optimal number of cluster.
Although, the mds plot maybe in this situation is not the best tool to check wheteher our clustering is good: we see some overlapping region and this might due to the fact that the mds lose information present in the original dataset.




#### Point b
In the second point we approach the problem in a different way: we generate an MDS and from the points generated we apply some different clustering techniques.

Let's take the points generated from a mds on two dimension.
```{r}
library(smacof)
point_mds<- mds(distmatrix, ndim=2)$conf
summary(point_mds)
```

# K-means
Let's apply the k-means algorithm to the dataset for different number of clusters

```{r}
library(fpc)
for (i in 2:15){
  mds_kmeans <- kmeans(point_mds, centers = i, iter.max = 100)
  summary(mds_kmeans)
  plot(point_mds,col=mds_kmeans$cluster,pch=clusym[mds_kmeans$cluster])
}

```
From a visual inspection we can see that when we have more than 10 clusters the point starts to overlap.


# Ward method
```{r}
clust_ward <- (hclust(distmatrix, method ="ward.D2"))
plot(clust_ward)

for (i in 2:15){
  ward_cut <- cutree(clust_ward, k = i)
  plot(point_mds, col = ward_cut, pch = clusym[ward_cut], )
}
```
Again, it's difficult to determine the optimal number of clusters, since there are some overlapping regions in almost every plot.


# Gaussian mixture
```{r}
mds_gaus_mixt <- Mclust(point_mds,G = 1:20)
summary(mds_gaus_mixt)
plot(mds_gaus_mixt)
```
---------------------------------------------------- 
Gaussian finite mixture model fitted by EM algorithm 
---------------------------------------------------- 

Mclust VVE (ellipsoidal, equal orientation) model with 7 components: 


Clustering table:
 1  2  3  4  5  6  7 
35 23 16 20 15 63 64 


Here we fitted a Gaussian mixture model to the mds points: the optimal mixture model estimation is VVE(ellipsoidal, equal orientation) with 7 components.


The nature of the problem involves the identification of bee species, aiming to discern various bee families. Despite a limited understanding of genomics, hierarchical clustering is suggested as a powerful tool. This method allows for the observation of similar traits gradually forming clusters, aiding in the grouping of families with shared characteristics.

If the bee species exhibit a more complex structure, yeilding irregular groupings, a non-hierarchical clustering could be a better tool.

Let's check whether to produce an MDS solution with more than two dimension could help to achieve to better results: higher dimensional spaces can capture more information but can also be more complex and harder to interpret.

## 3 dimension 

With the "rgl" library we can use the function $3dplot$ to see some visualization.

```{r}

tre_dim_mds <- mds(distmatrix, n = 3)$conf

x <- c(tre_dim_mds[,1])
y <- c(tre_dim_mds[,2])
z <- c(tre_dim_mds[,3])

point_mds_3d <- data.frame(x,y,z)
```


```{r}
mds_kmeans_3d <- kmeans(point_mds_3d, centers = 4)
plot3d(point_mds_3d, col=mds_kmeans_3d$cluster, type="s", size=3)
```



```{r}
mds_gaus_mixt_3d <- Mclust(point_mds_3d,G = 1:20)
summary(mds_gaus_mixt_3d)
plot(mds_gaus_mixt)
```

---------------------------------------------------- 
Gaussian finite mixture model fitted by EM algorithm 
---------------------------------------------------- 

Mclust VEV (ellipsoidal, equal shape) model with 8 components: 


Clustering table:
 1  2  3  4  5  6  7  8 
35 23 27 16 12 61 48 14 


With the 3-dimension mds, the optimal number of mixture components is 8 with VEV type. 



```{r}
pca_mixt_3d <- princomp(scale(point_mds_3d))

plot3d(pca_mixt_3d$scores,col=mds_gaus_mixt_3d$classification,
      pch=clusym[mds_gaus_mixt_3d$classification],
      type = "s",
      size = 2)
```






# Point 2


```{r}
wdbc <- read.csv("~/Desktop/Università/Unsupervised/wdbc.data", header=FALSE)
```

```{r}
wdbcc <- scale(wdbc[,3:12])
wdbcdiag <- as.factor(wdbc[,2])
```


```{r}
a <- kmeans(wdbcc, centers = 3)
plot(wdbcc,
     col=a$cluster,
     pch=clusym[a$cluster],
    )
```

```{r}
dist_breas <-dist(wdbcc, method = "euclidean")
mds_breast <- mds(dist_breas, ndim = 3)

x1 <- mds_breast$conf[,1]
x2 <- mds_breast$conf[,2]
x3 <- mds_breast$conf[,3]

mds_breast_dataframe <- data.frame(x1, x2, x3)

plot3d(mds_breast_dataframe,
       type = "s",
       size = 2,
       )

```


## Gaussian mixture model breast cancer
```{r}

wdbcc_gaus_mixt <- Mclust(wdbcc,G = 1:20)
summary(wdbcc_gaus_mixt)

```


```{r}
plot(wdbcc_gaus_mixt)
```
### which is better k-means or hierarchical? Probably, since the problem is to find different species based on similarity genes, I would use Hierarchical since you can see all the developement of the genes, so you can find familiy, races, species, ecc...

```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```








